---
title: "193DS_spring-2025_final"
author: "Mark Oregel"
date: "June 11, 2025"
format: html
toc: true 
toc-depth: 5
---

### GitHub repository: 
https://github.com/oregel714/ENVS-193DS_spring-2025_final

```{r, message=FALSE}

# reading in packages 
library(tidyverse)
library(here)
library(gt)
library(janitor)
library(readxl)
library(dplyr)
library(MuMIn)
library(DHARMa)
library(lubridate)
library(ggeffects)
library(fs)

# reading in data set 

sst <- read_csv(here("data", "SST_update2023.csv"))

nest_boxes <- read_csv(here("data", "occdist.csv"))

```




# Problem 1: 

### a. Transparents statistical methods:

In part 1, they used a Pearson's correlation test as this would mean that we are trying to understand if there is a correlation between distance from headwater and total nitrogen load. 
In part 2, they used an One-way Analysis of Variance (ANOVA) to determine if there is a difference between the sources of nitrogen and average nitrogen load. 

### b. More information needed: 

My co-worker can report the effect size of the one-way ANOVA test which would provide a measure of how much the grouping variable explains the response variable, in this case, the effect or magnitude of source type on average nitrogen load. This bolsters their results and provides the audience a measure of the difference between groups. Additionally, since their result is significant (p < 0.05), they can run a Tukey's Honestly Significant Difference (Tukey HSD) post-hoc test which would explain which groups are statistically significantly different when compared. For example, grassland and urban land sources might have significantly higher average nitrogen loads than fertilizer. This test would provide additional important information for understanding any significant differences between nitrogen sources and average nitrogen load. 

### c. Suggestions for rewriting: 

Part 1: We found a strong positive (Pearson's r = correlation coefficient) relationship between distance from headwater (km) and annual total nitrogen load (kg $year^-1$) where distance from headwater significantly predicted annual nitrogen load (p = 0.03). 

Part 2: We found a significant effect size ($r^2$ = effect size) between nitrogen sources in average nitrogen load (kg/$year^-1$) (one-way ANOVA, 'F' = distribution (among groups df, withing groups df) = F - statistic, p = 0.02, $a$ = significance level). On average, 'Tukey HSD would describe any statistically significant difference between pairs of groups' (a quantity greater or less than, followed by confidence interval [lower bound, upper bound] kg/$year^-1$) and repeated for the other group in comparison. 




# Problem 2: 

### a. Cleaning and Summarizing 

```{r, message=FALSE}

sst_clean <- sst |> # creating new data object from original data set
  select(date, temp) |> # selecting relevant 
  mutate( # changing columns to year and date 
    year = year(date), # specifiying dat as a year  
    month = month(date, label = TRUE, abbr = TRUE) # label and abbreviate the month date 
  ) |> 
  filter(year %in% c(2018, 2019, 2020, 2021, 2022, 2023)) |> # filtering the data for only the specified years 
  select(-date) |> # getting rid of the original date column 
  group_by(year, month) |> # grouping by year and month 
  summarise(mean_monthly_sst = mean(temp, na.rm = TRUE), .groups = "drop")  |> # summarizing  the mean monthly surface temperatures 
  mutate(mean_monthly_sst = round(mean_monthly_sst, 1), year = as.factor(year)) # changing the 'mean_monthly_sst' to be rounded and setting 'year' as a factor 

sst_clean |> # taking the 'sst_clean' object and sampling 5 observations 
  slice_sample(n=5)

str(sst_clean) # displaying the sample 
```

### b. Visualization 
```{r, message=FALSE}

ggplot(data = sst_clean, # starting with data set 'sst_clean' 
       aes(x = month, # setting x axis as 'month'
           y = mean_monthly_sst, # setting y axis as 'mean_monthly_sst'
           group = year, # grouping by year 
           color = year # coloring the year 
           )) +
  geom_line(linewidth = 0.5) + # setting the size of the line 
  geom_point(size = 2) + # setting the size of the points representing the means
  scale_color_manual( # setting the colors of each year manually 
    values = c(
      "2023" = "#292f56",
      "2022" = "#1e4572",
      "2021" = "#005c8b",
      "2020" = "#007498",
      "2019" = "#008ba0",
      "2018" = "#00a3a4"
    )
  ) +
  scale_x_discrete( # abbreviating the months on the x axis 
    limits = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  ) +
  
  theme_bw() + # setting the theme to black and white 
  labs(y = "Mean monthly sea surface temperature (°C)", # setting titles for axes 
       x = "Month", # setting title of x axis to month 
       color = "Year" # setting legend title to year 
       ) + 
  theme( # changing the size/position of the text in the legend and axes to fit the graph 
    legend.position = "inside",
    legend.position.inside = c(0.1, 0.75), 
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11),
    panel.grid.major = element_blank(), # removing the major grids/panels in the back of the graph
    panel.grid.minor = element_blank(), # removing the minor grids/panels 
    
  )
```

# Problem 3: 

```{r, message=FALSE}

clean_nb <- nest_boxes |> # creating new data object 
  clean_names() |> # cleaning names of columns 
  select(-box, -event_id, -repeated_use) |> # de-selecting 
  mutate(season = as.factor(season)) # indicating 'season' as a factor 

```


### a. Response Variable:

Biologically speaking, the 1s and 0s are binary values indicating which species is present in the nest box for the observation. There are three species, whichever species is present for that observation, a 1 denotes the presence of the species or an empty box while the other species are denoted by 0s to indicate their absence. 

### b. Purpose of Study: 

In the context of the study, nesting boxes are targeted towards the Swift Parrot species while the other species of interest, Common Starling and Tree Martin, are competitors who have been exploiting the nesting boxes more than the Swift Parrots. 

### c. Difference in "seasons":

The two seasons (years) the authors are referring to are 2016 and 2019. Subordinate native nest competitors increased the likelihood of box occupancy the farther away from the forest edge in both years, however, this relationship was observed to be stronger in 2019. 

### d. Table of models: 

4 models total:

| Model # | Season | Distance to forest edge   | Model description          |  
|:-------:|:------:|:-------------------------:|:--------------------------:|  
| 0       |        |                           | no predictors (null model) |
| 1       |    X   |            X              | all predictors (full model)| 
| 2       |    X   |                           | season                     | 
| 3       |        |            X              | distance to forest edge    |

### e. Run the models:

```{r, message=FALSE}

# model 0: null model 
model0 <- glm(
  sp ~ 1, # formula for null model 
  data = clean_nb, # data frame
  family = "binomial" # indicating distribution 
)

# model 1: saturated model 
model1 <- glm(
  sp ~ edge_distance + season,
  data = clean_nb,
  family = "binomial"
)

# model 2: season 
model2 <- glm(
  sp ~ season,
  data = clean_nb,
  family = "binomial"
)

# model 3: distance from forest edge 
model3 <- glm(
  sp ~ edge_distance,
  data = clean_nb,
  family = "binomial"
)

```

f. Check the diagnostics 

```{r, message=FALSE, model-diagnostics}

# plotting simulated residuals for all models 
plot(
  simulateResiduals(model0)
)
plot(
  simulateResiduals(model1)
)
plot(
  simulateResiduals(model2)
)
plot(
  simulateResiduals(model3)
)

```

### g. Select the best model: 

```{r model-selection, message=FALSE}

# using Akaike's Information Criterion to choose the best model 
AICc(
  model0,
  model1,
  model2,
  model3
) |> 
  arrange(AICc) # arranging in order from least to greatest descending 

```

The best model as determined by Akaike's Information Criterion (ACI) uses season and edge distance. 

### h. Visualize the model predictions: 

```{r visualizing the model predictions, message=FALSE}

mod_preds <- ggpredict(model1, # creating new data object from ggpredict function using model1
          terms = c("edge_distance [all]", "season")) # specifying terms and to include all data from edge_distance 

ggplot(clean_nb, # data frame 
       aes(x = edge_distance, # setting x axis as edge_distance 
           y = sp, # setting y axis as Swift Parrot 
           color = season)) + # setting color to season 
  geom_point(data = clean_nb, # plotting points using data from clean_nb
             size = 2, # setting size of points 
             alpha = 0.4) + # setting point transparency 
  geom_ribbon(data = mod_preds, # using data from mod_preds to create confidence interval ribbon 
              aes(x = x, # setting x axis as distance value from forest edge 
                  y = predicted, # setting y axis as predicted value 
                  ymin = conf.low, # setting bottom edge of the ribbon  
                  ymax = conf.high, # setting top edge of the ribbon 
                  fill = group), # filling based on group (season)
              alpha = 0.4, # adding transparency 
              color = NA) + # removing outline of ribbon 
  geom_line(data = mod_preds, # creating line using data from mod_preds
            aes(x = x, # setting x axis as edge_distance 
                y = predicted, # setting y axis as predicted 
                color = group), # assigning color by group (season)
            linewidth = 1) + # changing size 
  scale_y_continuous(limits = c(0, 1), # setting y axis with a limit of 0 to 1 
                     breaks = c(0, 1)) + # setting y axis to only label 0 to 1 
  theme_bw() + # setting color theme for elements 
  theme( 
    panel.grid.major = element_blank(), # removing panels 
    panel.grid.minor = element_blank() # removing panels 
  ) + 
  labs(x = "Distance from the Forest Edge (meters)", # setting custom titles for x axis 
       y = "Swift Parrot Occupancy Probability", # setting custom titles for y axis 
       color = "Year", # legend title 
       fill = "Year" # combining both legends into one 
       
  ) + 
  scale_color_manual(values = c("2016" = "olivedrab", # setting outline color manually 
                                "2019" = "blue")) + 
  scale_fill_manual(values = c("2016" = "olivedrab", # setting fill colors 
                               "2019" = "blue"))

```

### i. Write a caption for your figure: 

**Figure 1: The probability of Swift Parrot occupancy in a nesting box tends to decrease the farther away it is located from the forest edge.**  Data from "Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird" (Stojanovic et. al, October 28, 2021; https://doi.org/10.5061/dryad.83bk3j9sb). Data points represent nesting box occupancy probability for Swift Parrots as either occupied (1) or unoccupied (0) with color representing the season (year) in which they occur, either 2016 (green, n = 29) or 2019 (blue, n = 20). The solid lines represent predicted probabilities resulting from a binomial logistic regression model. The shaded ribbons surrounding the lines represent a 95% confidence interval around the predictions. The two different colors represent our two seasons of interest, 2016 and 2019. 

### j. Calculate the model predictions: 


```{r calculating model1 predictions, message=FALSE}
# model predictions for Swift Parrot occupancy probability with 95% confidence at 0 meters in 2016
ggpredict(model1,
          terms = c("edge_distance [0]", "season [2016]"))
# model predictions for Swift Parrot occupancy probability with 95% confidence at 900 meters in 2016
ggpredict(model1,
          terms = c("edge_distance [900]", "season [2016]"))
# model predictions for Swift Parrot occupancy probability with 95% confidence at 0 meters in 2019
ggpredict(model1,
          terms = c("edge_distance [0]", "season [2019]"))
# model predictions for Swift Parrot occupancy probability with 95% confidence at 900 meters in 2019
ggpredict(model1,
          terms = c("edge_distance [900]", "season [2019]"))
```

### k. Interpret your results: 

Based on the visualization in part h and the model predictions in part j, there is a negative relationship between distance from forest edge and probability of Swift Parrot occupancy in nesting boxes for both 2016 and 2019 seasons.

**Model Predictions for 2016:**
If the distance from the forest edge is 0 meters, the probability of a Swift Parrot occupying a nesting box is 0.48 (95% CI [0.33,0.64]). Additionally, if the distance from the forest edge is 900 meters, the probability of a Swift Parrot occupying a nesting box is 0.12 (95% CI [0.06,0.24]).

**Model Predictions for 2019:**
If the distance from the forest edge is 0 meters, the probability of a Swift Parrot occupying a nesting box is 0.30 (95% CI [0.18,0.44]). If the distance from the forest edge is 900 meters, the probability of a Swift Parrot occupying a nesting box is 0.06 (95% CI [0.03,0.13]). 

Biologically speaking, the probability of Swift Parrots occupying nesting boxes far away from the forest edge decreases due to competitors, such as Tree Martins, exploiting the nesting boxes in higher frequency as distance from the edge of the forest increased.  This implies that Tree Martins displayed a positive relationship with probability of nesting box occupancy and distance from the forest edge.


# Problem 4: 

### a. Comparing visualizations:

The visualization that I made for homework 2 was simple and did not display anything meaningful about my data and the patterns that were present. Granted this was when I had very few observations, but my point still stands. My visualization for homework 3 had means between groups, standard error, and quantiles expressing what I had hoped to gather from my personal observations. 

A similarity that I found is that I chose the same groups that I represented in homework 3 in homework 2 as well. Additionally, I decided to stay with 'money spent' as one of my variables for my visualizations in both homework 2 and 3. 

There aren't very many patterns in my homework 2 visualization other than that I classified my days where I am on campus longer than 6 hours as my long days and days less than 6 hours as my short days. There are more obvious patterns in my homework 3 visualization as this is where I was able to fully display my observations. I noticed that I typically spend more money on long days and on short days, I have a noticeable amount of observations where I spend no money at all. This makes perfect sense as I do my best to not spend money on days where I am on campus for a short amount of time, but alas, these observations would say otherwise. 

The feedback I received in week 9 in workshop primarily pertained to my artist statement and my use of drawings. People recommended that I implement a physical aspect to the visualization as my drawings would be able to be replicated onto the physical objects themselves; these being a yerba mate can and a coffee cup. I ended up drawing the points on these physical objects with a marker to make it stand out, unfortunately, my roommate mistook them as trash on our dining room table and threw them away :(. 

### b. Sharing your affective visualization

I attended workshop :). 









